---
title: "Functional Data Regression"
author: "Madhur Bansal (210572)"
date: "2024-03-14"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

# Simulating the data

We construct a random variable $X_{(t)}$ from $L^2_{[0, 1]}$ space using the following: $$X_{(t)} = (c_1 * e^t) + sin(c_2 * 10t) + (c_3*2t) $$ where $c_1$, $c_2$, $c_3$ are random variables drawn from $uniform(0, 2)$. We generate (n = 300) samples of $X_{(t)}$ as our given data. Here is the plot of 5 samples from the generated data.

```{r data_generation}
library(fda)

set.seed(1)

n_samples <- 300
n_test <- 100
n_time_pts <- 100

# Evaluates a function given the random coeff.
evaluate_x <- function(coeff, t)
{
  exp_1 <- coeff[1] * exp(t)
  exp_2 <- sin(coeff[2]* 10 * t)
  exp_3 <- coeff[3] * 2 * t
  
  return(exp_3 + exp_2 + exp_1)
}

# Random coeff for the given data
X_sample <- matrix(data = runif(3*n_samples, min = 0, max = 2),
            nrow = n_samples, byrow = T)

X_test <- matrix(data = runif(3*n_test, min = 0, max = 2),
                 nrow = n_test, byrow = T)

# Given Xi's plot
grid_x <- seq(0, 1, length = n_time_pts)
plot(x = 0, y = 0, xlim = c(0, 1), ylim = c(-1, 12), 
     xlab = "t", ylab = "X(t)",
     main = "Generated Data (5 samples)", type = 'n')

for(i in 1:5)
{
  grid_y <- evaluate_x(coeff = X_sample[i, ], t = grid_x)
  lines(grid_x, grid_y, col = i)
}
legend("topleft", legend = paste(1:5), 
       col = 1:5, lty = 1)
```

We simulate real $Y$ using the following:

$$ Y = m(X_{(t)}) + \epsilon $$

where $m(X_{(t)}) = \int_0^1X_{(t)}^2 (sin(t) + cos(t)) \space dt$ and $\epsilon$ \~ $N(0, 1)$

```{r calculating_y}
operator_x <- function(coeff)
{
  fn <- integrate(f = function(t){evaluate_x(coeff, t)^2 * (sin(t) + cos(t))},
                  lower = 0, upper = 1)$value
  return(fn)
}

y_sample <- numeric(length = n_samples)
y_sample <- apply(X = X_sample, MARGIN = 1, FUN = operator_x)
y_sample <- y_sample + rnorm(n = n_samples)

y_real <- apply(X = X_test, MARGIN = 1, FUN = operator_x)
```

Here are the real values of $y_i$ for above $X_{(t)}^i$ in respective order:

```{r print_y}
print(head(y_sample))
```

# Functional Linear Regression

Initially I predict the response variable using in-built functional linear regression function *fdRegress(.)*. The only reason to predict using the Linear Regression model is to get a baseline for model comparison. It tries to find $\beta$ such that:

$$ Y_i = \beta_0 + \int_0^1X_{i}(t)\beta(t)dt + \epsilon_i$$
Both $X_i(t)$ and $\beta(t)$ can be expanded using orthonormal basis functions in $L_{2}$ space, say K:

$$X_i(t) \approx \sum_{k=1}^Kc_{ik}\phi_k \space,  \space \beta(t)  \approx \sum_{k=1}^Kb_{k}\phi_k \space$$
Expanding the respective terms, we get:
$$Y_i = \beta_0 + \sum_{k = 1}^Kb_kc_{ik}$$

# Proposed Estimator (Based on Nadaraya-Watson Estimator)

I am using something similar to Nadaraya-Watson estimator for estimating $m$. The proposed estimator is as follows:

$$ \hat{m}(X_{(t)}) = \frac{\Sigma_{i = 1}^n K(\space ||X - X_{(t)}^i||_{L_2}\space / \space h) y_i}{\Sigma_{i = 1}^n K(\space ||X - X_{(t)}^i||_{L_2}/ h)}$$

where:

$K(.)$ is a valid kernel. I have used standard Normal $N(0, 1)$

$h$ is the band-width

$||.||$ is the norm on $L_{2[0, 1]}$ space

## Choosing the Bandwidth (h)

In order to choose the appropriate band-width for the simulated sample data, I am using K-fold cross validation.

Here are the steps I followed:

1.  I generate a grid of band-width containing 21 values - $h = \{0.05, 0.075, ... 0.55\}$

2.  Then I perform K-fold cross validation (with $K = 4$), and calculate the sum of the MSE obtained from cross-validation, for every h in the grid.

3.  Chose the $h$ which gives the lowest sum of MSE from the K cross validation results.

```{r choose_bandwidth, fig.align = 'center', fig.width = 4}
# x <- predict for this x
# bw <- bandwidth
# X_train <- given sample
# y_train <- given sample
predict_x <- function(x, bw, X_sample, y_sample)
{
  n <- length(y_sample)
  weights <- numeric(length = n)
  for(i in 1:n)
  {
    distance <- integrate(f = function(t){(evaluate_x(X_sample[i, ], t) - evaluate_x(x, t))^2},
                          lower = 0, upper = 1)$value
    distance <- sqrt(distance)
    weights[i] <- dnorm(x = distance/bw)
  }
  estimate <- weighted.mean(x = y_sample, w = weights)
  
  return(estimate)
}

# Cross Validation
k_cross_val <- function(X_sample, y_sample, bw, k = 4)
{
  # Stores mse for k-folds
  mse <- numeric(length = k)
  
  n <- length(y_sample)
  div <- n / k
  
  for(i in 1:k)
  {
    # Splitting
    X_train <- X_sample[-( ((i-1)*div + 1):(i*div) ), ]
    X_test <- X_sample[((i-1)*div + 1):(i*div), ]
    
    y_train <- y_sample[-( ((i-1)*div + 1):(i*div) )]
    y_test <- y_sample[((i-1)*div + 1):(i*div)]
    
    y_pred <- apply(X = X_test, MARGIN = 1, 
                    FUN = function(x){predict_x(x, bw, X_train, y_train)})
    
    mse[i] <- mean((y_test - y_pred)^2)
  }
  
  return(mse)
}

# Making a grid
grid_bw <- seq(from = 0.05, to = 0.55, length = 21)
mse_bw <- numeric(length = length(grid_bw))
for(i in 1:length(grid_bw))
{
  mse_bw[i] <- sum(k_cross_val(X_sample = X_sample, y_sample = y_sample,
                          bw = grid_bw[i], k = 4))
  # print(paste(i, "/", length(grid_bw), "done"))
}

plot(x = grid_bw, y = mse_bw,
     type = 'l', lwd = 2,
     xlab = "h", ylab = "sum of MSE",
     main = "sum of MSE vs Band-width")

chosen_bw <- grid_bw[which.min(mse_bw)]

print(paste("Chosen Band-width:", chosen_bw))
```

# Estimation for new data

Now, we generate 100 new $(X_{(t)}, \space y)$ in order to evaluate the performance of our estimator.

Here are some of the estimates made for the newly generated data:

## Estimation using Linear Regression:
```{r prediction_linear}

X_data_train <- matrix(0, nrow = n_samples, ncol = n_time_pts)
X_data_test <- matrix(0, nrow = n_test, ncol = n_time_pts)
for(i in 1:n_samples)
{
  X_data_train[i, ] <- evaluate_x(X_sample[i, ], grid_x)
}
for(i in 1:n_test)
{
  X_data_test[i, ] <- evaluate_x(X_test[i, ], grid_x)
}

y_data_train <- y_sample
y_data_test <- y_real

basis <- create.bspline.basis(rangeval = c(0, 1), nbasis = 10)
betalist <- list()
betalist[[1]] <- fdPar(basis, Lfdobj = 0, lambda = 1e-2)

fd_train <- Data2fd(argvals = grid_x, y = t(X_data_train), basisobj = basis)
fd_test <- Data2fd(argvals = grid_x, y = t(X_data_test), basisobj = basis)

fd_model <- fRegress(y_data_train, fd_train, betalist = betalist, method = "fRegress")
y_pred_lin <- predict(fd_model, newdata = list(fd_test))

mse_lin <- mean((y_real - y_pred_lin)^2)

predict_table <- cbind(y_real, y_pred_lin, y_real - y_pred_lin)
colnames(predict_table) = c("Real y", "Estimated y", "Error")

print(head(predict_table))

print(paste("MSE = ", mse_lin))

```

## Estimation using proposed estimator

```{r prediction_test}
y_pred_pro <- apply(X = X_test, MARGIN = 1, 
                FUN = function(x){predict_x(x, chosen_bw, X_sample, y_sample)})

mse_pro <- mean((y_real - y_pred_pro)^2)

predict_table <- cbind(y_real, y_pred_pro, y_real - y_pred_pro)
colnames(predict_table) = c("Real y", "Estimated y", "Error")

print(head(predict_table))

print(paste("MSE = ", mse_pro))
```

# Conclusion

Both estimators successfully predicted the dependent variable; however, the Nadaraya-Watson estimator outperformed the linear model slightly. This indicates that the linear model may not be the most appropriate fit for the data.

<!-- # Effect of sample size (n_samples) -->

<!-- Here we try to evaluate the performance of the estimator with changing value of n_samples. We will evaluate the same simulated for n_samples = {40, 100, 200, 300, 500} -->